Computational methods in landscape ecology

### 2024-09-18 Sections/03-connectivity.qmd
WIP: Shorten connectiviy section

Landscape connectivity describes how landscapes facilitate or impede movement among resource patches [@Taylor1993ConnectivityVital; @Taylor2006LandscapeConnectivity].
It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
Thus, multiple approaches how to measure connectivity were developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics].
Connectivity can be divided into structural and functional connectivity.
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

## Estimating connectivity

<!-- MH: Potential shorten here -->

While connectivity is key to many modern conservation theories and practices, quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high-resolution possible [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals for the required tempo-spatial resolutions.
Thus, in practice connectivity is primarily determined by indirect estimation.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of the physical attributes of the landscape and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

### Structural Connectivity Estimation Methods

Landscape metrics, produced using software such as `landscapemetrics`, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size. [@Spanowicz2019MeasuringLandscape].

### Estimating Functional Connectivity Using Resistance Surfaces

Functional connectivity estimation involves integrating species-specific movement data with landscape features.
This estimation normally involves the usage of graph-theoretic approaches, though these details are often obscured for convenience by software [@McRae2008UsingCircuit].
Two primary base approaches are usually applied in modern connectivity estimation, namely least-cost paths (LCP) and circuit theory.

The methods all rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell [@Etherington2016LeastcostModelling].
The values are based on landscape features and represent factors influencing movement, such as habitat type, topography, or barriers.

Creating a resistance surface involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
Next, the cost value for each cell need to be quantified using empirical movement data.
This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (Genetic samples used to calculate genetic distances between populations).

@@

Resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

#### Least-cost paths (LCP)

LCP models assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
By calculating the LCP between two points using a resistance surface, researchers can estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
The creation of LCPs is a well established technique and there are highly optimised and efficient algorithms available for its calculation.

#### Circuit theory

Circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
In this approach, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
Key measures from circuit theory are the resistance distance and the current flow.
The former can describes isolation by measuring multiple low- or high resistance pathways, the later identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape

>>>>

Landscape connectivity describes how landscape featuers facilitate or impede movement [@Taylor2006LandscapeConnectivity].
It underpins ecosystem functionality, maintains biodiversity and populations, and plays a large role in many conservation actions [@CorreaAyram2016HabitatConnectivity; @Hilty2020GuidelinesConserving].
Connectivity can be divided into structural and functional connectivity.
Structural connectivity describes the physical arrangement of landscape elements, focusing on spatial aspects such as continuity and adjacency.
It is solely a landscape feature and independent of species' characteristics [@Tischendorf2000UsageMeasurement; @CorreaAyram2016HabitatConnectivity].
Functional connectivity, in contrast, integrates landscape structure with the perceptual, behavioural, and dispersal characteristics of species, making it both species- and landscape-specific [@Adriaensen2003ApplicationLeastcost; @Taylor2006LandscapeConnectivity].

Multiple approaches how to measure connectivity have been developed [@Simpkins2018AssessingPerformance; @Keeley2021ConnectivityMetrics], however, simultaneously quantifying it has proven both technically and conceptually challenging [@Calabrese2004ComparisonshoppersGuide; @Taylor2006LandscapeConnectivity].
Data from many individuals recording the start location, trajectory, and endpoint of movement is required to infer the elements in the landscape that individuals preferentially move through [@Simpkins2018AssessingPerformance].
While improving technologies have made tracking animal movement at high tempo-spatial resolution available [@Kays2015TerrestrialAnimal; Nathan2022BigdataApproaches], it is still logistically challenging to track high numbers of individuals.
Thus, in practice connectivity is primarily determined by indirect estimations.
@Calabrese2004ComparisonshoppersGuide distinguished three types of estimates: i) structural connectivity determined by the physical attributes of the landscape, ii) potential connectivity as a combination of physical landscape attributes and limited information about dispersal capabilities of species, and iii) actual connectivity related to observations of individuals moving through a landscape.

### Estimating structural connectivity

Landscape metrics, are commonly used to quantify various aspects of landscape structure decomposed into composition and configuration.
These metrics describe structural connectivity in terms of total habitat amount, patch size and isolation, or possible connections between habitat patches.
Additionally, more recently there has been an increasing focus on within-patch connectivity using metrics such as the effective mesh size. [@Spanowicz2019MeasuringLandscape].

### Estimating functional connectivity

#### Resistance surfaces

Functional connectivity estimation involves integrating species-specific movement data with landscape features.
Many modern connectivity estimation approaches rely on resistance surfaces to represent the landscape [@Zeller2012EstimatingLandscape].
A resistance surface is a grid-based representation of a landscape where each cell is assigned a value representing the difficulty (or cost) for an individual of a particular species to traverse that cell based on landscape features such as habitat type, topography, or barriers [@Etherington2016LeastcostModelling].

Creating resistance surfaces involves obtaining landscape data for the area of interest along with a sufficiently large buffer area to avoid edge artifacts [@Dutta2022OverviewComputational].
Next, cost values for each cell need to be quantified using movement data.
This can included expert opinion [widely used due to low cost, however, difficult to measure accuracy\; @Krueger2012RoleExpert], detection data (single point locations of unknown individuals), relocation data (multiple sequential locations of the same individual but at low frequency), pathway data (high-frequency relocation data allowing for movement track inference), or genetic data (samples used to calculate genetic distances between populations).

@@

However, resistance surfaces are not free of critique including missing spatial-temporal variability or context dependency [@UnnithanKumar2022ConnectivityModelling; @UnnithanKumar2022MovingLandscape].
Thus, recent developments include the combination with more processes-based modelling approaches [@Schumaker2018HexSimModeling; @Bocedi2021RangeShifter20; @UnnithanKumar2022PathwalkerNew], modelling of dynamic landscape connectivity [@Zeller2020UnderstandingImportance], inclusion of stochasticity and spatial context [@Ribeiro2017LandScapeCorridors], using genetic optimization algorithms [@Peterman2018ResistanceGAPackage], or computational improvements [@Leonard2017GlowSoftware].

Least-cost paths assume that individuals have perfect knowledge of the landscape and will choose the path that minimizes resistance costs perfectly with no deviations [@Adriaensen2003ApplicationLeastcost].
By calculating paths between two points, it is possible to estimate connectivity between these points based on the accumulated cost along the path.
This method can also generate accumulated cost surfaces depicting the minimum cost from a single point to all other locations, helping identify reachable areas within a threshold.
The creation of least-cost paths is a well established technique and there are highly optimised and efficient algorithms available for its calculation.

<!-- This is missing some refs -->
<!-- Move to next section -->
Contrastingly, circuit theory considers all possible pathways between locations simultaneously [@McRae2008UsingCircuit].
In this approach, the landscape is represented as a network where nodes are connected by resistors (weighted by the values in the resistance surface), and movement is analogous to electrical current flow.
Circuit theory does not assume that organisms have perfect knowledge of the landscape and is particularly useful when multiple alternative pathways are available.
Key measures from circuit theory are the resistance distance and the current flow.
The former can describes isolation by measuring multiple low- or high resistance pathways, the later identifies areas of high movement probability, highlighting important corridors and bottlenecks in the landscape
